# Clonal Evolution analysis using ClonEvol

The following uses outputs from PyClone-VI to build a phylogenetic tree of the evolutionary pattern of clonal populaitons
in the chroangioma from the PVP project. This follows the instructions set forth by the ClonEvol authors in their
setup and use guide.

### Load in libraries
```{r message=FALSE, warning=FALSE}
install.packages("devtools", repos = "https://mirror.las.iastate.edu/CRAN/")
library(devtools)
install_github("hdng/clonevol")
install_github("hdng/trees")
install.packages("gridBase")
install.packages("gridExtra")
install.packages("ggplot2")
install.packages("igraph", type = "binary")
install.packages("packcircles")
```

## Workflow

From the authors docs the workflow of mapping the clonal evolution are:

  1. Preparing a comprehensive and reliable set of somatic mutations
  2. Clustering the mutations based on cellular prevalence
  3. Evaluating the clustering results
  4. Inferring clonal evolution trees (ie. clonal ordering)
  5. Visualizing data and clonal evolution trees
  6. Interpreting the results

Steps 1 and 2 are already already completed using Mutect2, CNVkit, and PyClone-VI. ClonEvol will be helping with
steps 3 thorugh 6.

## Analysis

### Data Prep

After reviewing the parameters for PyClone-VI and the outputs we'll be analyzing the clusters produced from the binomial
distribution. The authors of ClonEvol note: ClonEvol requires an input data frame consisting of at least a cluster
column and one or more variant cellular prevalence columns, each corresponds to a sample.
The cluster should be named by contiguous integer numbers, starting from 1. For better
visualization, the names of the cellular prevalence columns should be short.

Note from PyClone-VI authors: the column labeled "cellular_prevalence" = Proportion of malignant cells with the 
mutation in the sample. This is also called cancer cell fraction (CCF) in the literature.

```{r}
library(clonevol)
#df <- read.table(file = "data/pyclone-output/pvp-pyclonevi-binomial-clonal-prediction.tsv", header = TRUE, sep = "\t")
df <- read.table(file = "data/pyclone-output/pvp-pyclonevi-betabi-clonal-prediction.tsv", header = TRUE, sep = "\t")
```


ClonEvol expects cluster IDs to be integer values starting at 1, however PyClone-VI starts cluster IDs at zero so we
need to increase each cluster ID by 1 before we can generate inference. Guessing from the ClonEvol demo data we also
need to convert the CCF info from a decimal fraction to a percent, and possibly include depth information :shrug:

```{r}
df$cluster_id <- df$cluster_id + 1
df$cpp <- df$cellular_prevalence * 100

# convert dataframe to state where cellular prevalence is a column for each sample
# and drop unused columns
data_list <- split(df, f = df$sample_id)
df <- df[c("mutation_id", "cluster_id")]
sids <- names(data_list)
for (sid in sids) {
  df[[sid]] <- data_list[[sid]]$cpp[match(df$mutation_id, data_list[[sid]]$mutation_id)]
}
```

Another modification is required before processing b/c despite saying they can take other column names for the cluster ID
a known and unfixed issue has been logged https://github.com/hdng/clonevol/issues/40 with the exact problem I initially
ran into. Hence we need to rename the `cluster_id` column to `cluster`, the expected default name, before doing analysis.

```{r}
colnames(df)[which(names(df) == "cluster_id")] <- "cluster"
```

Another unfortunate bug was found working with ClonEvol, thank goodness for FOSS and issue tracking on GitHub!
This issue occurs when trying to run in single sample mode, the author of the tool notes that at some point they
stopped supporting a single sample mode explicitly and suggest just making a fake duplicate sample to satisfy the
tool and run the analysis. See https://github.com/hdng/clonevol/issues/38 for the full write up. The major interest
in this project is the clonal evolution of the chorangioma tissue so focus will simply be on that sample

```{r}
df$D <- df$LW001299
```

## Ordering and Tree building

Since we used PyClone-VI to estimate copy number corrected CCF values we can provide the corrected CCF to ClonEvol via 
`ccf.col.names` parameter in `infer.clonal.models` function:

```{r}
# set ClonEvol to the cluster colors desired, NULL indicates that ClonEvol will auto pick colors for us
clone.colors <- c("#f3722c","#43aa8b","#f9c74f","#577590")

# fix cluster ordering for the model so founding clone is always labeled as 1
# see https://github.com/hdng/clonevol/issues/7#issuecomment-382786214
c3 = df$cluster == 4
c1 = df$cluster == 1
df$cluster[c1] = 4
df$cluster[c3] = 1
df = df[order(df$cluster),]

y <- infer.clonal.models(
  variants = df,
  cluster.col.name = "cluster",
  ccf.col.names = c("LW001299", "D"),
  cancer.initiation.model = "monoclonal",
  subclonal.test = "bootstrap",
  subclonal.test.model = "non-parametric",
  num.boots = 1000,
  founding.cluster = 1,
  cluster.center = "mean",
  ignore.clusters = NULL,
  clone.colors = clone.colors,
  min.cluster.vaf = 0.01,
  # min probability that CCF(clone) is non-negative
  sum.p = 0.05,
  # alpha level in confidence interval estimate for CCF(clone)
  alpha = 0.05
)

y <- convert.consensus.tree.clone.to.branch(y, branch.scale = "log2")
```

Plot the evolution dynamics in a PDF

```{r}
plot.clonal.models(
  y,
  disable.sample.label=TRUE,
  # box plot parameters
  box.plot = TRUE,
  fancy.boxplot = TRUE,
  fancy.variant.boxplot.jitter.alpha = 1,
  fancy.variant.boxplot.jitter.center.color = "grey50",
  fancy.variant.boxplot.base_size = 12,
  fancy.variant.boxplot.plot.margin = 1,
  fancy.variant.boxplot.vaf.suffix = " Cellular Fraction",
  # bell plot parameters
  clone.shape = "bell",
  bell.event = TRUE,
  bell.event.label.color = "blue",
  bell.event.label.angle = 60,
  clone.time.step.scale = 1.25,
  bell.curve.step = 1,
  # node-based consensus tree parameters
  merged.tree.plot = FALSE,
  tree.node.label.split.character = NULL,
  tree.node.shape = "circle",
  tree.node.size = 30,
  tree.node.text.size = 0.5,
  merged.tree.node.size.scale = 1.25,
  merged.tree.node.text.size.scale = 1.5,
  merged.tree.cell.frac.ci = FALSE,
  # branch-based consensus tree parameters
  merged.tree.clone.as.branch = TRUE,
  mtcab.event.sep.char = ',',
  mtcab.branch.text.size = 1,
  mtcab.branch.width = 0.25,
  mtcab.node.size = 6,
  mtcab.node.label.size = 2,
  mtcab.node.text.size = 1,
  # cellular population parameters
  cell.plot = TRUE,
  num.cells = 200,
  cell.border.size = 0.25,
  cell.border.color = "black",
  clone.grouping = "horizontal",
  # meta-parameters
  scale.monoclonal.cell.frac = TRUE,
  show.score = FALSE,
  cell.frac.ci = TRUE,
  disable.cell.frac = FALSE,
  # output figure parameters
  out.dir = "data/clonevol-output",
  out.format = "pdf",
  overwrite.output = TRUE,
  width = 16,
  height = 10,
  # vector of width scales for each panel from left to right
  panel.widths = c(3, 5, 4, 4)
)

print("Done")
```